### 2025-07-26 paper/camera/appendices/additional.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\section{Additional Experiments}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/main-mlp.pdf}
    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios with the MLP architecture.
    We show the mean and standard deviation of the normalized rewards across $3$ runs. For the MLP architecture see \Cref{fig:experiments}.
    }\vspace{0.5cm}
    \label{apx:fig:experiments-mlp}

\end{figure}

>>>>

\section{Additional Experiments}\label{apx:experiments}

@@

To check also the scalability of \fname{} we run the experiments with up to $20$ agents in the Transport scenario. The results are shown in \Cref{apx:fig:experiments-mlp}. We observe that \fname{} is able to scale well with the number of agents, achieving good results even with $20$ agents, also in this case similarly to SHAC.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/main-mlp.pdf}
    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios with the MLP architecture.
    We show the mean and standard deviation of the normalized rewards across $3$ runs. For the MLP architecture see \Cref{fig:experiments}.
    }\vspace{0.5cm}
    \label{apx:fig:experiments-mlp}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/transport-high-agents.pdf}
    \caption{Comparison between \fname{} and SHAC for increasing number of agents (up to 20) in the Transport scenario with the Transformer architecture. Results demonstrate the scalability of both approaches as the number of agents increases.}\label{apx:fig:experiments-mlp-20}

    \vspace{0.5cm}
\end{figure}

### 2025-07-26 paper/camera/sects/abstract.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\begin{abstract}
Reinforcement learning (RL) algorithms show promise in robotics and multi-agent systems but often suffer from low sample efficiency. While methods like SHAC leverage differentiable simulators to improve efficiency, they are limited to specific settings: they require fully differentiable environments (including transition and reward functions) and have primarily been demonstrated in single-agent scenarios. To overcome these limitations, we introduce \fname{}, a novel framework inspired by SHAC\@. \fname{} removes the need for differentiable simulator components by using neural networks to approximate the required gradients, training these networks alongside the standard policy and value networks. This allows the core SHAC approach to be applied in both non-differentiable and multi-agent environments. We evaluate \fname{} on challenging multi-agent tasks from the VMAS suite, comparing it against SHAC (where applicable) and PPO, a standard algorithm for non-differentiable settings. Our results demonstrate that \fname{} significantly outperforms PPO in both single- and multi-agent scenarios. Furthermore, in differentiable environments where SHAC operates, \fname{} achieves comparable performance despite lacking direct access to simulator gradients, thus successfully extending SHAC's benefits to a broader class of problems.
\end{abstract}

>>>>

\begin{abstract}
Reinforcement learning (RL) algorithms show promise in robotics and multi-agent systems but often suffer from low sample efficiency. While methods like SHAC leverage differentiable simulators to improve efficiency, they are limited to specific settings: they require fully differentiable environments, including transition and reward functions, and have primarily been demonstrated in single-agent scenarios. To overcome these limitations, we introduce \fname{}, a novel framework inspired by SHAC. \fname{} removes the need for differentiable simulator components by using neural networks to approximate the required gradients, training these networks alongside the standard policy and value networks. This enables the core SHAC approach to be applied in both non-differentiable and multi-agent environments. We evaluate \fname{} on challenging multi-agent tasks from the VMAS suite, comparing it against SHAC (where applicable) and PPO, a standard algorithm for non-differentiable settings. Our results demonstrate that \fname{} significantly outperforms PPO in both single- and multi-agent scenarios. Furthermore, in differentiable environments where SHAC operates, \fname{} achieves comparable performance despite lacking direct access to simulator gradients, thus successfully extending SHAC's benefits to a broader class of problems. The full implementation is openly available at 
\begin{center}
    \url{https://github.com/f14-bertolotti/shacpp}.
\end{center}
\end{abstract}

### 2025-07-26 paper/camera/sects/background.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. 
Particularly, we focus on multi-agent settings, where multiple agents interact with each other and the environment.
In doing so, each task is then modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$. Where $\agents=\{1, 2, \dots, \na\}=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\allo={\{\observations{i}\}}_{i\in\agents}$ and $\alla={\{\actions{i}\}}_{i\in\agents}$ are the observation space and action space for each agent. ${\{\stateobs{i}:\states\rightarrow\observations{i}\}}_{i\in\agents}$ are the projections from the state to the observation space and ${\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}}_{i\in\agents}$ are the reward functions for each agent. $\transition:\states\times\alla\rightarrow\states$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

@@

While we use a unique set of parameters $\theta$ for all the networks. Each network can either use its own subset of $\theta$ or it can share parameters with other networks. In practice, the parameters will be shared only for the policies.

For simplicity, we formally consider only the case where policies have access to the full state, meaning that $\stateobs{i}$ is the identity function. This allows us to focus on the fundamental aspects of our approach without the added complexity of partial observability. Extending our framework to handle partial observability is straightforward; however, we choose not to address it here to maintain clarity and simplicity in our presentation. Nonetheless, our experiments do include scenarios where policies do not have access to the complete state information.

@@

@@

$G^t_k$ represents the discounted sum of rewards up to step $k$, plus the bootstrapped value of the state at step $t+k$. The estimates $V$ is computed by discounting longer (higher $k$) discounted sums, $G^k_t$.

>>>>

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning framework. 
Particularly, we focus on multi-agent settings, where multiple agents interact with each other and the environment.
In doing so, each task is then modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$. Where $\agents=\{1, 2, \dots, \na\}=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\allo={\{\observations{i}\}}_{i\in\agents}$ and $\alla={\{\actions{i}\}}_{i\in\agents}$ are the observation space and action space for each agent. ${\{\stateobs{i}:\states\rightarrow\observations{i}\}}_{i\in\agents}$ are the projections from the state to the observation space and ${\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}}_{i\in\agents}$ are the reward functions for each agent. $\transition:\states\times\alla\rightarrow\states$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

@@

Throughout this notation, we use a single global set of parameters $\theta$ to describe all networks. Each network can either use its own distinct subset of $\theta$ or share parameters with other networks. In practice, parameter sharing occurs only among the policy networks.

To keep the notation simple, we formally consider only the case where policies have access to the full state, meaning that $\stateobs{i}$ is the identity function. Extending our framework to handle partial observability is straightforward. Nonetheless, our experiments do include scenarios where policies do not have access to the complete state information.

@@

Short Horizon Actor-Critic (SHAC) is a model-based RL algorithm designed to leverage differentiable simulators for efficient policy learning. It alternates between rolling out short trajectories (a fixed small horizon h) in parallel and updating an actor–critic pair using gradients that flow directly through the simulator’s dynamics and reward model.

@@

$G^k_t$ represents the discounted sum of rewards up to step $k$, plus the bootstrapped value of the state at step $t+k$. The estimates $V$ is computed by discounting longer (higher $k$) discounted sums, $G^k_t$.

### 2025-07-26 paper/camera/sects/experiments.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.9\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Policy gradient norms in the Transport environment over 5000 epochs. Spikes in gradient norm for SHAC (epoch 2700) and \fname{} (epoch 3200) correspond to generalization events when agents begin frequent interactions with the package and each other.}\label{fig:grads-transformer-transport}
    \vspace{0.5cm}
\end{figure}
\section{Experiments}\label{sect:experiments}

In this section, we evaluate \fname{} against both PPO and SHAC across various multi-agent scenarios and architectures. An ablation study replacing only the transition function (SHAC+, where the reward function remains differentiable) is provided in \Cref{apx:ablation}.
We select PPO as our baseline because, despite the emergence of more recent algorithms HAPPO \cite{Kuba2021TrustRM}, and SAC \cite{Haarnoja2018SoftAA}, it remains one of the few approaches that scales effectively to multi-agent settings without substantial modifications (e.g., MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22}). This makes PPO a representative benchmark for state-of-the-art performance in non-differentiable environments.

@@

\begin{table*}[t]
    \centering
    \input tables/max.tex
    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
\end{table*}

@@

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds. See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters.

@@

\paragraph{Hardware Setup}
Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one NVIDIA Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).

\subsection{Results}
\begin{figure}
    \centering
    \includegraphics[width=0.32\columnwidth]{figs/step-1.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-2.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-3.png}
    \caption{Snapshots of trained agents with \fname{} in the Transport scenario. Colored circles represent agents pushing a red package (square) toward the green goal.}\label{fig:transport-trained}
    \vspace{0.5cm}
\end{figure}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

@@

>>>>

\section{Experiments}\label{sect:experiments}

In this section, we evaluate \fname{} against both PPO (for single agent settings), MAPPO (for multi-agent settings), and SHAC across various multi-agent scenarios and architectures. An ablation study replacing only the transition function (SHAC+, where the reward function remains differentiable) is provided in \Cref{apx:ablation}.
We select PPO as our baseline because, despite the emergence of more recent algorithms HAPPO \cite{Kuba2021TrustRM}, and SAC \cite{Haarnoja2018SoftAA}, it remains one of the few approaches that scales effectively to multi-agent settings without substantial modifications (e.g., MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22}). This makes PPO a representative benchmark for state-of-the-art performance in non-differentiable environments.

@@

@@

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds. See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters. Hyperparameter choices are the result of a small grid search on the transport scenario.

@@

\paragraph{Hardware Setup}
All experiments were conducted on a cluster with two nodes, each equipped with a V100 GPU (32GB of memory) and 200GB of RAM, and one node with four A100 GPUs (40GB of memory) and 512GB of RAM.

\begin{figure}
    \centering
    \includegraphics[width=0.32\columnwidth]{figs/step-1.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-2.png}
    \includegraphics[width=0.32\columnwidth]{figs/step-3.png}
    \caption{Snapshots of trained agents with \fname{} in the Transport scenario. Colored circles represent agents pushing a red package (square) toward the green goal.}\label{fig:transport-trained}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Policy gradient norms in the Transport environment over 5000 epochs. Spikes in gradient norm for SHAC (epoch 2700) and \fname{} (epoch 3200) correspond to generalization events when agents begin frequent interactions with the package and each other.}\label{fig:grads-transformer-transport}
    \vspace{0.5cm}
\end{figure}

\subsection{Results}

In this section, we compare the performance of \fname{}, PPO/MAPPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO/MAPPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
\end{figure*}

\begin{table*}
    \centering
    \input tables/max.tex
    \vspace{0.2cm}
    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
\end{table*}

### 2025-07-26 paper/camera/sects/introduction.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\section{Introduction}\label{sect:introduction}
Learning control policies is crucial for applications like robotics \cite{Singh22}, autonomous driving \cite{Elallid22}, and swarm intelligence \cite{Tang21}. These domains often present challenges such as sparse rewards, complex dynamics, and multi-agent interactions. Many current RL approaches focus specifically on either single- or multi-agent settings, lacking a unified solution. Furthermore, RL algorithms are often \emph{sample inefficient}, requiring extensive interaction with the environment to learn effective policies.

A promising direction to improve sample efficiency involves utilizing \emph{differentiable simulators}, namely simulators that allow for backpropagation through their components. SHAC \cite{Xu22} exemplifies this approach, using gradients obtained directly from the simulator to accelerate policy optimization. However, SHAC's applicability is constrained by two key limitations. First, it necessitates both a differentiable transition function \emph{and} a differentiable reward function, which is often impractical, especially with sparse or complex reward structures. Second, while powerful, backpropagation through complex dynamics, like object collisions~\cite{Georgiev24} or intricate multi-agent interactions, can lead to unstable gradients~\cite{Bengio94, Metz21}, potentially hindering learning. The original SHAC work also focused primarily on single-agent tasks.

In this work, we propose \fname{} to address these limitations and extend gradient-based policy optimization beyond SHAC's constraints. Our goal is to develop a sample-efficient reinforcement learning framework that maintains the benefits of gradient-based optimization without requiring differentiable simulators. Drawing inspiration from model-based RL \cite{DBLP:conf/icml/HafnerLFVHLD19, DBLP:conf/iclr/HafnerLB020}, \fname{} substitutes direct simulator differentiation with neural approximations of the gradient functions. By training networks that approximate transition and reward gradients concurrently with policy optimization, our approach preserves the sample efficiency benefits of gradient information while eliminating strict differentiability requirements. This design enhances robustness to complex dynamics and enables application in both non-differentiable environments and multi-agent scenarios, albeit with additional computational overhead compared to SHAC.

We evaluate \fname{} against two established baselines: PPO \cite{Schulman17} (including its multi-agent variant MAPPO \cite{DBLP:conf/nips/YuVVGWBW22}) as the standard for non-differentiable environments, and the original SHAC \cite{Xu22} where applicable. Our experiments utilize the VMAS simulator \cite{DBLP:conf/dars/BettiniKBP22}, which provides physics-based multi-agent environments with varying levels of complexity. Through this comparative analysis on cooperative tasks with different agent populations, we investigate the following research questions:

\begin{compactitem}
    \item \RQ{1} Can we train a neural network to approximate the gradients of a differentiable simulator?
    \item \RQ{2} How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?
    \item \RQ{3} How does the performance of these algorithms change as the search space increases?
\end{compactitem}

>>>>

\section{Introduction}\label{sect:introduction}
Learning control policies is crucial for applications such as robotics \cite{Singh22}, autonomous driving \cite{Elallid22}, and swarm intelligence \cite{Tang21}. These domains often present challenges including sparse rewards, complex dynamics, and multi-agent interactions. Many current RL approaches focus specifically on either single- or multi-agent settings, lacking a unified solution. Furthermore, RL algorithms are often \emph{sample inefficient}, requiring extensive interaction with the environment to learn effective policies.

A promising direction to improve sample efficiency involves utilizing \emph{differentiable simulators}, namely simulators that allow backpropagation through their components. SHAC \cite{Xu22} exemplifies this approach by using gradients obtained directly from the simulator to accelerate policy optimization. However, SHAC’s applicability is constrained by two key limitations. First, it necessitates both a differentiable transition function \emph{and} a differentiable reward function, which is often impractical, especially in the presence of sparse or complex reward structures. Second, while powerful, backpropagation through complex dynamics, such as object collisions~\cite{Georgiev24} or intricate multi-agent interactions, can lead to unstable gradients~\cite{Bengio94, Metz21}, potentially hindering learning. The original SHAC work also focused primarily on single-agent tasks.

In this work, we propose \fname{} to address these limitations and extend gradient-based policy optimization beyond SHAC's constraints. Our goal is to develop a sample-efficient reinforcement learning framework that maintains the benefits of gradient-based optimization without requiring differentiable simulators. Drawing inspiration from model-based RL \cite{DBLP:conf/icml/HafnerLFVHLD19, DBLP:conf/iclr/HafnerLB020}, \fname{} substitutes direct simulator differentiation with neural approximations by training networks that approximate the transition and reward functions concurrently with the policy function. This approach preserves the benefits of gradient information while eliminating strict differentiability requirements. It also enhances robustness to complex dynamics and enables application in both non-differentiable environments and multi-agent scenarios, albeit with additional computational overhead compared to SHAC.

We evaluate \fname{} against two established baselines: PPO \cite{Schulman17} (including its multi-agent variant MAPPO \cite{DBLP:conf/nips/YuVVGWBW22}), as the standard for non-differentiable environments, and the original SHAC \cite{Xu22}, where applicable. Our experiments utilize the VMAS simulator \cite{DBLP:conf/dars/BettiniKBP22}, which provides physics-based multi-agent environments with varying levels of complexity. Through this comparative analysis on cooperative tasks with different agent populations, we investigate the following research questions:

\begin{compactitem}
    \item \RQ{1} Can we train a neural network to approximate the gradients of a differentiable simulator?
    \item \RQ{2} How does our algorithm compare to PPO/MAPPO and SHAC in both single-agent and multi-agent settings?
    \item \RQ{3} How does the performance of these algorithms change as the search space increases?
\end{compactitem}

### 2025-07-26 paper/camera/sects/related.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\paragraph{Deep Reinforcement Learning}
Deep reinforcement learning has revolutionized policy learning tasks by combining reinforcement learning with deep neural networks.
Foundational works like DQN~\cite{Mnih13} (and its variants~\cite{DBLP:conf/aaai/HesselMHSODHPAS18})
and AlphaZero~\cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess.
Over the years, several algorithms have been proposed to create more stable and efficient learning methods.
Notable examples include A3C for asynchronous learning~\cite{DBLP:conf/icml/MnihBMGLHSK16},
DDPG and PPO for continuous action spaces~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17} (also used in modern chatbot like ChatGPT),
and SAC for continuous action spaces with off-policy learning~\cite{DBLP:conf/icml/HaarnojaZAL18}.
Following the initial era of model-free algorithms,
model-based algorithms have been introduced to enhance sample efficiency.
Early works such as Dreamer~\cite{DBLP:conf/iclr/HafnerLB020} and PlaNet~\cite{DBLP:conf/icml/HafnerLFVHLD19} have demonstrated promising results in terms of both sample efficiency and performance.

\paragraph{Multi-Agent Reinforcement Learning}
Multi-Agent Reinforcement Learning (MARL) is a subfield of reinforcement learning that focuses on learning policies for multiple agents that interact with each other~\cite{albrecht2024multi}.
This field has a wide range of applications,
including robotics~\cite{DBLP:journals/air/ChungFYN24},
traffic control~\cite{DBLP:journals/tits/ChuWCL20},
and game playing~\cite{DBLP:conf/iclr/BakerKMWPMM20,jaderberg2019human},
and involves different types of agents such as cooperative, competitive, and mixed.
In this paper, we primarily focus on cooperative multi-agent scenarios,
where agents need to collaborate to achieve a common goal.
Prominent works in this area include extensions of single-agent algorithms to multi-agent settings,
such as MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17} and MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22},
as well as novel solutions to unique challenges in multi-agent settings. Examples of these solutions are COMA~\cite{DBLP:conf/aaai/FoersterFANW18} for multi-agent credit assignment,
QMIX~\cite{DBLP:conf/icml/RashidSWFFW18} for coordination in cooperative multi-agent settings,
and Mean Field MARL~\cite{DBLP:conf/icml/YangLLZZW18} for handling a large number of agents.

>>>>

\paragraph{Deep Reinforcement Learning}
Deep reinforcement learning has revolutionized policy learning tasks by combining reinforcement learning with deep neural networks.
Foundational works like DQN~\cite{Mnih13} (and its variants~\cite{Hessel18})
and AlphaZero~\cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess.
Over the years, several algorithms have been proposed to create more stable and efficient learning methods.
Notable examples include A3C for asynchronous learning~\cite{Mnih16},
DDPG and PPO for continuous action spaces~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17} (also used in modern chatbot like ChatGPT),
and SAC for continuous action spaces with off-policy learning~\cite{DBLP:conf/icml/HaarnojaZAL18}.
Following the initial era of model-free algorithms,
model-based algorithms have been introduced to enhance sample efficiency.
Early works such as Dreamer~\cite{DBLP:conf/iclr/HafnerLB020} and PlaNet~\cite{DBLP:conf/icml/HafnerLFVHLD19} have demonstrated promising results in terms of both sample efficiency and performance.

\paragraph{Multi-Agent Reinforcement Learning}
Multi-Agent Reinforcement Learning (MARL) is a subfield of reinforcement learning that focuses on learning policies for multiple agents that interact with each other~\cite{albrecht2024multi}.
This field has a wide range of applications,
including robotics~\cite{DBLP:journals/air/ChungFYN24},
traffic control~\cite{DBLP:journals/tits/ChuWCL20},
and game playing~\cite{DBLP:conf/iclr/BakerKMWPMM20,jaderberg2019human},
and involves different types of agents such as cooperative, competitive, and mixed.
In this paper, we primarily focus on cooperative multi-agent scenarios,
where agents need to collaborate to achieve a common goal.
Prominent works in this area include extensions of single-agent algorithms to multi-agent settings,
such as MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17} and MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22},
as well as novel solutions to unique challenges in multi-agent settings. Examples of these solutions are COMA~\cite{DBLP:conf/aaai/FoersterFANW18} for multi-agent credit assignment,
QMIX~\cite{DBLP:conf/icml/RashidSWFFW18} for coordination in cooperative multi-agent settings,
and Mean Field MARL~\cite{Yang18} for handling a large number of agents.

### 2025-07-26 paper/camera/sects/shacpp.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

Similarly, we aim to train a neural network to approximate the transition function. However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}. This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $\alla^\mh$, predicts the next $\mh$ states from $\states^\mh$. We aim to minimize:
\begin{align*}
    \underset{\tau\sim\pi}{E}\left[\hspace{-1.2cm}\sum_{\scriptscriptstyle \hspace{1cm}\substack{s_t,\dots,s_{t+h+1},\\\hspace{.3cm}a_{t*},\dots,a_{t+h*}\in\tau}}\hspace{-1.2cm} \left\vert\left\vert \transition_{\theta}(s_t, a_{t*},\dots,a_{\scriptscriptstyle i+h*}) - [s_t,\dots,s_{\scriptscriptstyle t+h+1}] \right\vert\right\vert^2\right] 
\end{align*}
Where $a_{t*}=\{a_{ti}\}_{i\in\agents}$. In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, a_{1*}^\tau, \dots, a_{\mh*}) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
Where $B$ represents a batch of trajectories of length $\mh$ sampled uniformly from the cache.

Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space. Suppose the action space is $(l,r)$ (with $l<r$) for all agents, we minimize the loss $\mathcal{L'}_\theta^\pi:$

@@

    Where $\alpha$ is a hyper-parameter that controls the strength of the penalty. In our experiments, we set $\alpha=1$. Usually, actions are either clipped to the allowed action space or, an activation function (such as hyperbolic tangent or sigmoid) is used to ensure that the actions are within the allowed range. While this approach is acceptable for algorithms like PPO and SHAC, the penalty we employ is crucial in our specific scenario. This necessity arises from the fact that we are training the policy to optimize two neural networks that serve as environment approximations. Because these networks are never trained with data outside the permissible action space, they are likely to be maximized beyond these bounds. Consequently, the policy may generate actions far outside the allowed range. This can ultimately result in 0-gradients when action clipping is applied or high pre-activation actions when activation function is used.

@@

The full implementation is openly available at \url{https://anonymous.4open.science/r/diffagents-8653}\footnote{The link of the original repository will be provided upon acceptance.}.

>>>>

Similarly, we aim to train a neural network to approximate the transition function. However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}. This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $\alla^\mh$, predicts the next $\mh$ states from $\states^\mh$. We aim to minimize:
\begin{align*}
    \underset{\tau\sim\pi}{E}\left[\hspace{-1.2cm}\sum_{\scriptscriptstyle \hspace{1cm}\substack{s_t,\dots,s_{t+h+1},\\\hspace{.3cm}a_{t*},\dots,a_{t+h*}\in\tau}}\hspace{-1.2cm} \left\vert\left\vert \transition_{\theta}(s_t, a_{t*},\dots,a_{\scriptscriptstyle i+h*}) - [s_t,\dots,s_{\scriptscriptstyle t+h+1}] \right\vert\right\vert^2\right] 
\end{align*}
Where $a_{t*}$ denotes the set of actions taken from all agents at time $t$, $\{a_{ti}\}_{i\in\agents}$. In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, a_{1*}^\tau, \dots, a_{\mh*}) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
Where $B$ represents a batch of trajectories of length $\mh$ sampled uniformly from the cache.

Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space. Suppose the action space is the open interval $(l,r)$ for all agents, we minimize the loss $\mathcal{L'}_\theta^\pi:$

@@

    Here, $\alpha$ is a hyperparameter that controls the strength of the penalty, and in our experiments we set $\alpha=1$. Typically, actions are either clipped to the allowed action space or constrained via activation functions such as hyperbolic tangent or sigmoid to ensure they remain within valid bounds. While these approaches are sufficient for algorithms like PPO and SHAC, in our case the penalty is crucial. This is because our policy is trained to optimize two neural networks that approximate the environment dynamics and reward. Since these networks are only trained on data within the valid action space, their outputs are unreliable for out-of-bound actions. As a result, the policy may exploit these approximations by generating actions far outside the allowed range. Action clipping may lead to zero gradients, while activation functions may results large pre-activation values.

@@
